---
title: "Probe and Sample Filtering"
output: html_document
---

```{r library, message=F, warning=F}
library(minfi)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(sas7bdat)
library(openxlsx)
 

```

# EARLI Cord Processing

## Load preprocessed data and covariates

```{r}
res.path <- "/dcl01/NDEpi/data/Projects/InProgress/jdou/DNAm_EARLI_MARBLES/EARLI_cord/"

#load noob data
load("/dcl01/NDEpi/data/Projects/InProgress/kbakulsk/kbakulsk/EARLI/450k-round2/Noob-beta-All-Samples-EARLI-both-rounds.rda")

#load pd
load("/dcl01/NDEpi/data/MasterCohortData/EARLI/EARLIBiologicalSamples/450k/EARLI_450k_all1155samples_pd_with_ancestry.rda")
load("/dcl01/NDEpi/data/MasterCohortData/EARLI/EARLIBiologicalSamples/450k/earli.450k.pd.N=1225.20161117.rda")

interview <- read.csv("/dcl01/NDEpi/data/MasterCohortData/EARLI/EARLISelfReportedInterviewData/Data Extraction Codebook 20150203_export_20150203_1422987480_wide_20150204_1423070501.csv")
covars <- read.sas7bdat("/dcl01/NDEpi/data/Projects/InProgress/jdou/MARBLES/e_m_covars_v5_20191115.sas7bdat")

```

## Sample and Probe QC filtering

```{r}
### samp dropping

#get the cord blood
pd.cord <- pd[pd$Tissue=='Cord.Blood',]
pd.cord <- merge(pd.cord,pd.ancest[,c('Tissue.Cat','predictedSex')],by='row.names',sort=FALSE)
rownames(pd.cord) <- pd.cord$Row.names
nrow(pd.cord) #175

#remove multiple births
multi <- pd.cord[duplicated(pd.cord$DCC_Family_ID),"DCC_Family_ID"]
pd.cord <- pd.cord[!pd.cord$DCC_Family_ID %in% multi,]
nrow(pd.cord) #173

#drop those with discordant predicted and observed sex
discordant.sex <- pd.cord[!is.na(pd.cord$Sex) & pd.cord$Sex != pd.cord$predictedSex,]
nrow(discordant.sex) # 1
pd.cord <- pd.cord[!rownames(pd.cord)%in%rownames(discordant.sex),]
nrow(pd.cord) #170



### probe dropping

#drop poor detection probes
load('/dcl01/NDEpi/data/Projects/InProgress/kbakulsk/kbakulsk/EARLI/450k-round2/Failed-Probes-By-DetP-01.rda')
pd.cord$detp_fail_pct <- colMeans(probe.fail[,rownames(pd.cord)])
bad.probe <- rowMeans(probe.fail)
bad.probe <- bad.probe[bad.probe>0.05]
length(bad.probe) #661 failed probes
noob.beta <- noob.beta[!(rownames(noob.beta)%in%names(bad.probe)),]
rm(bad.probe)
nrow(noob.beta) #484851

#drop cross reactive
cross <- read.csv("/dcl01/NDEpi/data/Projects/InProgress/jdou/48639-non-specific-probes-Illumina450k.csv", stringsAsFactors=FALSE)
noob.beta <- noob.beta[!rownames(noob.beta)%in%cross$TargetID,]
rm(cross)
nrow(noob.beta) #455698

#drop Y probes
sex <- Locations[Locations$chr %in% c('chrY'), ]
noob.beta <- noob.beta[!rownames(noob.beta) %in% rownames(sex), ]
dim(noob.beta) #455650



#select samps
noob.beta <- noob.beta[,rownames(pd.cord)]


#extract interview data for samples
interview.fam <- interview[interview$Family_ID %in% pd.cord$DCC_Family_ID,]
questions <- c('EARLI_HB_smoke')
interview.q <- interview.fam[,c('Family_ID',questions)]
interview.q <- interview.q[match(pd.cord$DCC_Family_ID, interview.q$Family_ID),]

table(pd.cord$Subject.ID %in% covars$coi_id)
covars <- covars[match(pd.cord$Subject.ID, covars$coi_id), ]


#drop X probes
sex <- Locations[Locations$chr %in% c('chrX','chrY'), ]
beta.glob <- noob.beta[!rownames(noob.beta) %in% rownames(sex), ]
dim(beta.glob) #445241

### create globalDNAm 
source("/dcl01/NDEpi/data/Projects/InProgress/jdou/Folate/EARLI_folate_intake_functions.R")

pd.cord$meanDNAm <- colMeans(beta.glob)*100
pd.cord$meanDNAm.sea <- colMeans(genomic.region(beta.glob,region='OpenSea',anno='450k'))*100
pd.cord$meanDNAm.shore <- colMeans(genomic.region(beta.glob,region='Shore',anno='450k'))*100
pd.cord$meanDNAm.shelf <- colMeans(genomic.region(beta.glob,region='Shelf',anno='450k'))*100
pd.cord$meanDNAm.island <- colMeans(genomic.region(beta.glob,region='Island',anno='450k'))*100


```

## Get covariates formatted

```{r covar format}
#set up data frame with model variables
pd.cov <- data.frame(COI_ID = pd.cord$Subject.ID)
rownames(pd.cov) <- pd.cord$Row.names

#asd
pd.cov$asd <- covars$Dx_alg

#mat edu
pd.cov$mom_edu <- covars$MomEdu
pd.cov$mom_edu_bn <- ifelse(pd.cov$mom_edu<4, "No Degree", "College Degree")
pd.cov$mom_edu_bn <- factor(pd.cov$mom_edu_bn, levels=c("No Degree","College Degree"))

#maternal age
pd.cov$mom_age <- covars$MomAgeYr

#gest age
pd.cov$gest_age <- covars$GA_deliv_wks

#mat smk
pd.cov$smk_preg <- interview.q$EARLI_HB_smoke
pd.cov$smk_preg <- factor(pd.cov$smk_preg)

#ancestry
pd.cov$ancest_cat <- pd.cord$popid.questionnaire
pcs <- pd.cord[,which(colnames(pd.cord)=='PC1'):which(colnames(pd.cord)=='PC10')]
pd.cov <- cbind(pd.cov, pcs)

#sex
pd.cov$sex <- pd.cord$predictedSex

#cell type
cell.est <- readRDS("/dcl01/NDEpi/data/Projects/InProgress/jdou/Cell_Est/earl_cell_est_combi_idol.rds")
cell.est <- cell.est[rownames(pd.cov),]
identical(rownames(pd.cov), rownames(cell.est))
pd.cov <- cbind(pd.cov,cell.est)

#global DNAm
pd.cov$meanDNAm <- pd.cord$meanDNAm
pd.cov$meanDNAm.sea <- pd.cord$meanDNAm.sea
pd.cov$meanDNAm.shore <- pd.cord$meanDNAm.shore
pd.cov$meanDNAm.shelf <- pd.cord$meanDNAm.shelf
pd.cov$meanDNAm.island <- pd.cord$meanDNAm.island


saveRDS(pd.cov, file.path(res.path,'QC/pd.rds'))
saveRDS(noob.beta, file.path(res.path,'QC/beta.rds'))
```



# EARLI Placenta Processing

## Load preprocessed data and covariates

```{r}
res.path <- "/dcl01/NDEpi/data/Projects/InProgress/jdou/DNAm_EARLI_MARBLES/EARLI_placenta/"

#load noob data
load("/dcl01/NDEpi/data/Projects/InProgress/kbakulsk/kbakulsk/EARLI/450k-round2/Noob-beta-All-Samples-EARLI-both-rounds.rda")

#load pd
load("/dcl01/NDEpi/data/MasterCohortData/EARLI/EARLIBiologicalSamples/450k/EARLI_450k_all1155samples_pd_with_ancestry.rda")
load("/dcl01/NDEpi/data/MasterCohortData/EARLI/EARLIBiologicalSamples/450k/earli.450k.pd.N=1225.20161117.rda")

interview <- read.csv("/dcl01/NDEpi/data/MasterCohortData/EARLI/EARLISelfReportedInterviewData/Data Extraction Codebook 20150203_export_20150203_1422987480_wide_20150204_1423070501.csv")
covars <- read.sas7bdat("/dcl01/NDEpi/data/Projects/InProgress/jdou/MARBLES/e_m_covars_v5_20191115.sas7bdat")

```

## Sample and Probe QC filtering

```{r}
### samp dropping

#placenta
pd.placenta <- pd[pd$Tissue=='placenta',]
pd.placenta$predictedSex <- pd.ancest[rownames(pd.placenta),'predictedSex']
pd.plc <- pd.placenta[pd.placenta$Phenotype=='placenta.child',]
nrow(pd.plc) #134

#remove multiple births
multi <- pd.plc[duplicated(pd.plc$DCC_Family_ID),"DCC_Family_ID"]
pd.plc <- pd.plc[!pd.plc$DCC_Family_ID %in% multi,]
nrow(pd.plc) #128

#drop those with discordant predicted and observed sex
discordant.sex <- pd.plc[!is.na(pd.plc$Sex) & pd.plc$Sex != pd.plc$predictedSex,]
nrow(discordant.sex) # 1
pd.plc <- pd.plc[!rownames(pd.plc)%in%rownames(discordant.sex),]
nrow(pd.plc) #127



### probe dropping

#drop poor detection probes
load('/dcl01/NDEpi/data/Projects/InProgress/kbakulsk/kbakulsk/EARLI/450k-round2/Failed-Probes-By-DetP-01.rda')
pd.plc$detp_fail_pct <- colMeans(probe.fail[,rownames(pd.plc)])
bad.probe <- rowMeans(probe.fail)
rm(probe.fail)
bad.probe <- bad.probe[bad.probe>0.05]
length(bad.probe) #661 failed probes
noob.beta <- noob.beta[!(rownames(noob.beta)%in%names(bad.probe)),]
nrow(noob.beta) #484851

#drop cross reactive
cross <- read.csv("/dcl01/NDEpi/data/Projects/InProgress/jdou/48639-non-specific-probes-Illumina450k.csv", stringsAsFactors=FALSE)
noob.beta <- noob.beta[!rownames(noob.beta)%in%cross$TargetID,]
rm(cross)
nrow(noob.beta) #455698

#drop sex probes
sex <- Locations[Locations$chr %in% c('chrY'), ]
noob.beta <- noob.beta[!rownames(noob.beta) %in% rownames(sex), ]
dim(noob.beta) #455650   


#select samps
noob.beta <- noob.beta[,rownames(pd.plc)]

rm(multi, pd, pd.ancest, pd.placenta)



#extract interview data for samples
interview.fam <- interview[interview$Family_ID %in% pd.plc$DCC_Family_ID,]
questions <- c('EARLI_HB_smoke')
interview.q <- interview.fam[,c('Family_ID',questions)]
interview.q <- interview.q[match(pd.plc$DCC_Family_ID, interview.q$Family_ID),]

table(pd.plc$Subject.ID %in% covars$coi_id)
covars <- covars[match(pd.plc$Subject.ID, covars$coi_id), ]


#drop sex probes
sex <- Locations[Locations$chr %in% c('chrX'), ]
beta.glob <- noob.beta[!rownames(noob.beta) %in% rownames(sex), ]
dim(beta.glob)  


### create globalDNAm 
source("/dcl01/NDEpi/data/Projects/InProgress/jdou/Folate/EARLI_folate_intake_functions.R")

pd.plc$meanDNAm <- colMeans(beta.glob)*100
pd.plc$meanDNAm.sea <- colMeans(genomic.region(beta.glob,region='OpenSea',anno='450k'))*100
pd.plc$meanDNAm.shore <- colMeans(genomic.region(beta.glob,region='Shore',anno='450k'))*100
pd.plc$meanDNAm.shelf <- colMeans(genomic.region(beta.glob,region='Shelf',anno='450k'))*100
pd.plc$meanDNAm.island <- colMeans(genomic.region(beta.glob,region='Island',anno='450k'))*100
```

## Get covariates formatted

```{r covar format}
#set up data frame with model variables
pd.cov <- data.frame(COI_ID = pd.plc$Subject.ID)
rownames(pd.cov) <- pd.plc$Row.names

#asd
pd.cov$asd <- covars$Dx_alg

#mat edu
pd.cov$mom_edu <- covars$MomEdu
pd.cov$mom_edu_bn <- ifelse(pd.cov$mom_edu<4, "No Degree", "College Degree")
pd.cov$mom_edu_bn <- factor(pd.cov$mom_edu_bn, levels=c("No Degree","College Degree"))

#maternal age
pd.cov$mom_age <- covars$MomAgeYr

#gest age
pd.cov$gest_age <- covars$GA_deliv_wks

#mat smk
pd.cov$smk_preg <- interview.q$EARLI_HB_smoke
pd.cov$smk_preg <- factor(pd.cov$smk_preg)

#ancestry
pd.cov$ancest_cat <- pd.plc$popid.questionnaire
pcs <- pd.plc[,which(colnames(pd.plc)=='PC1'):which(colnames(pd.plc)=='PC10')]
pd.cov <- cbind(pd.cov, pcs)

#sex
pd.cov$sex <- pd.plc$Sex

#cell type
library(EpiDISH)
library(planet)
data('plCellCpGsThird')
out.l <- epidish(beta.m = noob.beta, ref.m = plCellCpGsThird, method = "RPC") 
cell.est <- out.l$estF
#saveRDS(cell.est, file="/dcl01/NDEpi/data/Projects/InProgress/jdou/Cell_Est/earl_placenta_epidish_planet.rds")
identical(rownames(pd.cov), rownames(cell.est))
pd.cov <- cbind(pd.cov,cell.est)

#global DNAm
pd.cov$meanDNAm <- pd.plc$meanDNAm
pd.cov$meanDNAm.sea <- pd.plc$meanDNAm.sea
pd.cov$meanDNAm.shore <- pd.plc$meanDNAm.shore
pd.cov$meanDNAm.shelf <- pd.plc$meanDNAm.shelf
pd.cov$meanDNAm.island <- pd.plc$meanDNAm.island


saveRDS(pd.cov, file.path(res.path,'QC/pd.rds'))
saveRDS(noob.beta, file.path(res.path,'QC/beta.rds'))
```



```{r libraryM, message=F, warning=F}
library(minfi)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(sas7bdat)
library(openxlsx)
library(dplyr)
```

# MARBLES  Cord Processing

## Load preprocessed data and covariates

```{r}
res.path <- "/dcl01/NDEpi/data/Projects/InProgress/jdou/DNAm_EARLI_MARBLES/MARBLES_cord/"

setwd("/dcl01/NDEpi/data/Projects/InProgress/jdou/MARBLES_EPIC_blood/QC/")
noob.beta <- readRDS("noob_beta.rds")
pd <- readRDS("qc_04_snp.rds")

detp <- readRDS("probe_fail_noY.rds")
cross <- read.csv("/dcl01/NDEpi/data/Projects/InProgress/jdou/cross_probes_info_EPIC_Pidsley_2016.csv")
flag <- read.xlsx('/dcl01/NDEpi/data/Projects/InProgress/jdou/MARBLES_EPIC_blood/flagged_samples.xlsx')
 
marb_cov <- read.sas7bdat("/dcl01/NDEpi/data/Projects/InProgress/jdou/MARBLES/marbles_allcov_09mar20.sas7bdat")
marbearl <- read.sas7bdat("/dcl01/NDEpi/data/Projects/InProgress/jdou/MARBLES/e_m_covars_v4.sas7bdat")

```


## Sample and Probe QC filtering

```{r}
### samp dropping

#this will axe controls
pd.an <- pd[pd$ew_probe_fail_pct<0.05, ]
nrow(pd.an) # 279

#drop methylation clock flagged samples
pd.an$flag_age <- pd.an$COI_ID %in% flag$ID
pd.an <- pd.an[!pd.an$flag_age,]
nrow(pd.an) # 247

#multiple births
table(pd.an$COI_ID %in% marb_cov$COI_ID)
fam <- marb_cov[match(pd.an$COI_ID, marb_cov$COI_ID), ]
multi <- fam[duplicated(fam$familyID),"familyID"]
multi <- multi[!is.na(multi)]
fam.multi <- fam[fam$familyID %in% multi, c("COI_ID","familyID","Year_DOB","PREGNUM")]
fam.multi <- fam.multi[order(fam.multi$familyID, fam.multi$PREGNUM),]
multiple.births <- c("5065","5389","5455","5456")
pd.an$FAM_ID <- substr(pd.an$COI_ID,1,4)
pd.an <- pd.an[!pd.an$FAM_ID %in% multiple.births, ]
nrow(pd.an) # 239

#sex mismatch gone
pd.an <- pd.an[is.na(pd.an$COI_GENDER) | !(pd.an$COI_GENDER=="M" & pd.an$predicted_sex=="f"), ]
nrow(pd.an) # 236

#keep one of related siblings
sib <- pd.an[duplicated(pd.an$FAM_ID),"FAM_ID"]
pd.an <- pd.an %>% group_by(FAM_ID) %>%
  filter(ew_probe_fail_pct == min(ew_probe_fail_pct))
pd.an <- data.frame(pd.an)
rownames(pd.an) <- pd.an$meth_id
nrow(pd.an) #223



### probe dropping

beta <- noob.beta[rownames(noob.beta) %in% rownames(Locations), ]
dim(beta) #865859

#drop detp fail probes
bad <- detp[detp>0.05]
beta <- beta[!rownames(beta) %in% names(bad), ]
dim(beta) #861229

#drop cross probes
beta <- beta[!rownames(beta) %in% cross$Probe, ]
dim(beta) #818262    286

#drop sex probes
sex <- Locations[Locations$chr %in% c('chrY'), ]
beta <- beta[!rownames(beta) %in% rownames(sex), ]
dim(beta) #817883    286


#select samps
beta <- beta[,match(rownames(pd.an), colnames(beta))]

rm(bad, cross, detp)


### select from covars data
cov <- marb_cov[match(pd.an$COI_ID, marb_cov$COI_ID),]




### create globalDNAm 
source("/dcl01/NDEpi/data/Projects/InProgress/jdou/Folate/EARLI_folate_intake_functions.R")

#drop X probes
sex <- Locations[Locations$chr %in% c('chrX'), ]
beta.glob <- beta[!rownames(beta) %in% rownames(sex), ]
dim(beta) #800191    286

pd.an$meanDNAm <- colMeans(beta.glob)*100
pd.an$meanDNAm.sea <- colMeans(genomic.region(beta.glob,region='OpenSea',anno='EPIC'))*100
pd.an$meanDNAm.shore <- colMeans(genomic.region(beta.glob,region='Shore',anno='EPIC'))*100
pd.an$meanDNAm.shelf <- colMeans(genomic.region(beta.glob,region='Shelf',anno='EPIC'))*100
pd.an$meanDNAm.island <- colMeans(genomic.region(beta.glob,region='Island',anno='EPIC'))*100
```


## Get covariates formatted

```{r covar format}
#set up data frame with model variables
pd.cov <- data.frame(COI_ID = pd.an$COI_ID)
rownames(pd.cov) <- pd.an$meth_id

#ASD
table(cov$Dx_alg, exclude=NULL)
table(cov$dxalg_flag)
pd.cov$dx_alg <- cov$Dx_alg
pd.cov$dx_flag <- cov$dxalg_flag

#maternal age
summary(cov$MomAgeYr)
pd.cov$mom_age <- cov$MomAgeYr
class(pd.cov$mom_age)

#gest age
summary(cov$GA_Deliv_wk)
pd.cov$gest_age <- cov$GA_Deliv_wk

#mat edu
table(cov$MomEdu, exclude=NULL)
pd.cov$mom_edu <- cov$MomEdu
pd.cov$mom_edu_bn <- ifelse(pd.cov$mom_edu<4, "No Degree", "College Degree")
pd.cov$mom_edu_bn <- factor(pd.cov$mom_edu_bn, levels=c("No Degree","College Degree"))

#mat smk
table(cov$SmokeYN_Pregnancy, exclude=NULL)
pd.cov$smk_preg <- cov$SmokeYN_Pregnancy
pd.cov$smk_preg <- factor(pd.cov$smk_preg)

#ancestry
table(cov$ChildRace, cov$ChildHisp, exclude=NULL)
pd.cov$race <- cov$ChildRace
pd.cov$eth <- cov$ChildHisp
pd.cov$ancest_cat <- ifelse(is.na(pd.cov$race), NA, 
                            ifelse(pd.cov$race==1 & pd.cov$eth==0, "euro", 
                                   ifelse(pd.cov$eth %in% c(1,2), "hisp", "other")))
pcs <- read.csv("/dcl01/NDEpi/data/Projects/InProgress/jdou/MARBLES/MARBLES-clean-chr1to22-GWAdata.pca.csv", header=T)
pcs$COI_ID <- substr(pcs$gid,1,6)
pcs <- pcs[match(pd.cov$COI_ID, pcs$COI_ID),]
pd.cov <- cbind(pd.cov, pcs[,2:11])

#sex
pd.cov$sex <- pd.an$COI_GENDER

#cell type
cell.est <- readRDS("/dcl01/NDEpi/data/Projects/InProgress/jdou/Cell_Est/marb_cell_est_combi_idol.rds")
cell.est <- cell.est[rownames(pd.cov),]
identical(rownames(pd.cov), rownames(cell.est))
pd.cov <- cbind(pd.cov,cell.est)

#global DNAm
pd.cov$meanDNAm <- pd.an$meanDNAm
pd.cov$meanDNAm.sea <- pd.an$meanDNAm.sea
pd.cov$meanDNAm.shore <- pd.an$meanDNAm.shore
pd.cov$meanDNAm.shelf <- pd.an$meanDNAm.shelf
pd.cov$meanDNAm.island <- pd.an$meanDNAm.island

saveRDS(pd.cov, file.path(res.path,'QC/pd.rds'))
saveRDS(beta, file.path(res.path,'QC/beta.rds'))
```


# MARBLES Placenta Processing

## Load preprocessed data and covariates

```{r}
res.path <- "/dcl01/NDEpi/data/Projects/InProgress/jdou/DNAm_EARLI_MARBLES/MARBLES_placenta/"

load("/dcl01/NDEpi/data/Projects/InProgress/jdou/EPIC_Placenta_Preprocess/pd.info.rda")
load("/dcl01/NDEpi/data/Projects/InProgress/jdou/EPIC_Placenta_Preprocess/untouched.noob.rda")

data(Locations)
load("/dcl01/NDEpi/data/Projects/InProgress/jdou/EPIC_Placenta_Preprocess/detP.min.rda")

cross <- read.csv("/dcl01/NDEpi/data/Projects/InProgress/jdou/cross_probes_info_EPIC_Pidsley_2016.csv")

marb_cov <- read.sas7bdat("/dcl01/NDEpi/data/Projects/InProgress/jdou/MARBLES/marbles_allcov_09mar20.sas7bdat")
```

## Sample and Probe QC filtering

```{r}
### samp dropping

#this will axe controls
pd.an <- pd.info
nrow(pd.an) # 92

detP.min <- detP.min[,rownames(pd.an)]
failedP <- detP.min > 0.01
pd.an$detp_fail_pct <- colMeans(failedP)
bad.probe <- rowMeans(failedP)
rm(detP.min, failedP)

#multiple births
table(pd.an$COI_ID %in% marb_cov$COI_ID)
fam <- marb_cov[match(pd.an$COI_ID, marb_cov$COI_ID), ]
multi <- fam[duplicated(fam$familyID),"familyID"]
multi <- multi[!is.na(multi)]
fam.multi <- fam[fam$familyID %in% multi, c("COI_ID","familyID","Year_DOB","PREGNUM")]
fam.multi <- fam.multi[order(fam.multi$familyID, fam.multi$PREGNUM),]
#none

#sex mismatch gone
table(pd.an$predictedSex, fam$COI_SEX)
#none

#keep one of related siblings
pd.an$FAM_ID <- substr(pd.an$COI_ID,1,4)
sib <- pd.an[duplicated(pd.an$FAM_ID),"FAM_ID"]
pd.an <- data.frame(pd.an)
pd.an <- pd.an %>% group_by(FAM_ID) %>%
  filter(detp_fail_pct == min(detp_fail_pct))
pd.an <- data.frame(pd.an)
rownames(pd.an) <- pd.an$Basename
nrow(pd.an) #90


### probe dropping

beta <- noob[rownames(noob) %in% rownames(Locations), ]
dim(beta) #865859

#detection p
bad <- bad.probe[bad.probe>0.05]
beta <- beta[!rownames(beta) %in% names(bad), ]
dim(beta) #864160

#drop cross probes
beta <- beta[!rownames(beta) %in% cross$Probe, ]
dim(beta) #821092

#drop sex probes
sex <- Locations[Locations$chr %in% c('chrY'), ]
beta <- beta[!rownames(beta) %in% rownames(sex), ]
dim(beta) #821008


#select samps
beta <- beta[,match(rownames(pd.an), colnames(beta))]
beta.noob <- getBeta(beta)
beta <- beta.noob
rm(beta.noob)

### select from covars data
cov <- marb_cov[match(pd.an$COI_ID, marb_cov$COI_ID),]


### create globalDNAm 
source("/dcl01/NDEpi/data/Projects/InProgress/jdou/Folate/EARLI_folate_intake_functions.R")

#drop X probes
sex <- Locations[Locations$chr %in% c('chrX'), ]
beta.glob <- beta[!rownames(beta) %in% rownames(sex), ]

pd.an$meanDNAm <- colMeans(beta.glob)*100
pd.an$meanDNAm.sea <- colMeans(genomic.region(beta.glob,region='OpenSea',anno='EPIC'))*100
pd.an$meanDNAm.shore <- colMeans(genomic.region(beta.glob,region='Shore',anno='EPIC'))*100
pd.an$meanDNAm.shelf <- colMeans(genomic.region(beta.glob,region='Shelf',anno='EPIC'))*100
pd.an$meanDNAm.island <- colMeans(genomic.region(beta.glob,region='Island',anno='EPIC'))*100
```

## Get covariates formatted

```{r covar format}
#set up data frame with model variables
pd.cov <- data.frame(COI_ID = pd.an$COI_ID)
rownames(pd.cov) <- pd.an$Basename

#ASD
table(cov$Dx_alg, exclude=NULL)
table(cov$dxalg_flag)
pd.cov$dx_alg <- cov$Dx_alg
pd.cov$dx_flag <- cov$dxalg_flag

#maternal age
summary(cov$MomAgeYr)
pd.cov$mom_age <- cov$MomAgeYr
class(pd.cov$mom_age)

#gest age
summary(cov$GA_Deliv_wk)
pd.cov$gest_age <- cov$GA_Deliv_wk

#mat edu
table(cov$MomEdu, exclude=NULL)
pd.cov$mom_edu <- cov$MomEdu
pd.cov$mom_edu_bn <- ifelse(pd.cov$mom_edu<4, "No Degree", "College Degree")
pd.cov$mom_edu_bn <- factor(pd.cov$mom_edu_bn, levels=c("No Degree","College Degree"))

#mat smk
table(cov$SmokeYN_Pregnancy, exclude=NULL)
pd.cov$smk_preg <- cov$SmokeYN_Pregnancy
pd.cov$smk_preg <- factor(pd.cov$smk_preg)

#ancestry
table(cov$ChildRace, cov$ChildHisp, exclude=NULL)
pd.cov$race <- cov$ChildRace
pd.cov$eth <- cov$ChildHisp
pd.cov$ancest_cat <- ifelse(is.na(pd.cov$race), NA, 
                            ifelse(pd.cov$race==1 & pd.cov$eth==0, "euro", 
                                   ifelse(pd.cov$eth %in% c(1,2), "hisp", "other")))
pcs <- read.csv("/dcl01/NDEpi/data/Projects/InProgress/jdou/MARBLES/MARBLES-clean-chr1to22-GWAdata.pca.csv", header=T)
pcs$COI_ID <- substr(pcs$gid,1,6)
pcs <- pcs[match(pd.cov$COI_ID, pcs$COI_ID),]
pd.cov <- cbind(pd.cov, pcs[,2:11])

#sex
pd.cov$sex <- pd.an$predictedSex

#cell type
library(EpiDISH)
library(planet)
data('plCellCpGsThird')
out.l <- epidish(beta.m = beta, ref.m = plCellCpGsThird, method = "RPC") 
cell.est <- out.l$estF
#saveRDS(cell.est, file="/dcl01/NDEpi/data/Projects/InProgress/jdou/Cell_Est/marb_placenta_epidish_planet.rds")
identical(rownames(pd.cov), rownames(cell.est))
cell.est <- cell.est[rownames(pd.cov),]
pd.cov <- cbind(pd.cov,cell.est)

#global DNAm
pd.cov$meanDNAm <- pd.an$meanDNAm
pd.cov$meanDNAm.sea <- pd.an$meanDNAm.sea
pd.cov$meanDNAm.shore <- pd.an$meanDNAm.shore
pd.cov$meanDNAm.shelf <- pd.an$meanDNAm.shelf
pd.cov$meanDNAm.island <- pd.an$meanDNAm.island


saveRDS(pd.cov, file.path(res.path,'QC/pd.rds'))
saveRDS(beta, file.path(res.path,'QC/beta.rds'))

pdf('planet_epidish_marbles_cell_est.pdf')
  hist(pd.cov$Trophoblasts)
  hist(pd.cov$Stromal)
  hist(pd.cov$Hofbauer)
  hist(pd.cov$Endothelial)
  hist(pd.cov$nRBC)
  hist(pd.cov$Syncytiotrophoblast)
dev.off()
```
